Comparing \pkg{binsegRcpp} with other implementations of binary segmentation for change-point detection

### 2025-05-13 article.Rnw
two changes fig

\section{Distributions and loss functions}

We assume there is a sequence of $N$ data $x_1,\dots,x_N\in\mathbb R$,
in which we want to find abrupt changes.  We
further assume that on each segment there is a constant model
parameter $\theta$, typically the mean (and perhaps
variance/scale). We would like to find a sequence of change-points,
such that each segment has a parameter $\theta$ which minimizes a loss
function $\ell(\theta,x)$. The loss functions that we consider to
minimize can be interpreted as maximizing the log likelihood of
several common distributions (Normal, Poisson, Laplace).

@@

Two models have loss functions that involve the absolute error
function, and the Laplace distribution. The probability density
function of the Laplace distribution, for data $x\in\mathbb R$, with
location parameter $\mu\in\mathbb R$, and scale parameter $b\in\mathbb
R$, is
\begin{equation}
  \frac{1}{2b} \exp\left(
  \frac{|\mu-x|}{-b}
  \right).
\end{equation}
For \verb|distribution.str="l1"|, the segment-specific parameter
$\theta=\mu\in\mathbb R$ is the median. In this case, maximizing the
probability corresponds to minimizing the negative log likelihood,
which simplifies to the absolute error loss function,
\begin{equation}
  \ell_{\texttt{l1}}(\mu, x) = |\mu-x|.
\end{equation}

@@

%% using the cumulative median
%% For loss functions which require the cumulative median, we use the C++ code below
%% The loss function which requires a variance estimate is the Laplace negative log likelihood; the simple L1 loss is obtained without a variance estimate.
A \code{class absDistribution} implements the common operations for
the \texttt{l1} and \texttt{laplace} loss functions, including
parameter estimation via a cumulative median algorithm, which is
required to efficiently compute the optimal split point on a segment.
In particular, when we search for the best split on a segment with
$s=e-j+1$ data, we need to compute the sequence of parameter estimates
$\mu^*_{j,j},\dots,\mu^*_{j,e-1}$ before the split, and
$\mu^*_{j+1,e},\dots,\mu^*_{e,e}$ after the split.  Computing these
parameter estimates and corresponding loss values can be done in
$O(s\log s)$ time, using a modified version of the L1 loss
minimization code proposed by \citet{Drouin2017mmit}.

@@

There are three models which use the cumulative sum trick to find the
best split for $N$ data in $O(N)$ time.
In particular, we define the vectors $W,X,Y\in\mathbb R^{N+1}$ as follows:
\begin{eqnarray}
  \text{Cumulative weights: } W_0=0,\,&& \forall t\in\{1,\dots,N\},\, W_t=\sum_{i=1}^t w_i,\\
  \text{Cumulative sum: }X_0=0,\,&& \forall t\in\{1,\dots,N\},\, X_t=\sum_{i=1}^t w_i x_i,\\
  \text{Cumulative squares: }Y_0=0,\,&& \forall t\in\{1,\dots,N\},\, Y_t=\sum_{i=1}^t w_i x^2_i.
\end{eqnarray}

@@

For \verb|distribution.str="mean_norm"|, the normal change in mean model,
the segment-specific parameter $\theta\in\mathbb R$ is the mean.
For data $x\in\mathbb R$, mean $\mu\in\mathbb R$, and variance $\sigma^2>0$, the normal probability density function is
\begin{equation}
  \frac{1}{\sqrt{2\pi\sigma^2}}
  \exp\left[
    \frac{
      (\mu-x)^2
    }{
      -2\sigma^2
    }
    \right].
\end{equation}
Maximizing the normal probability is equivalent to minimizing the negative log-likelihood, which simplifies to the square loss,
\begin{equation}
  \ell_{\texttt{mean\_norm}}(\mu, x) = (\mu-x)^2.
\end{equation}
For a segment from $j$ to $e$ ($1\leq j\leq e\leq N$), we use the notation $W_{j,e}=W_j-W_{e-1}$ for the sum of the weights over the segment, and similarly for $X_{j,e}$ and $Y_{j,e}$. Then the mean on the segment from $j$ to $e$ can be computed via
\begin{equation}
  \mu^* = X_{j,e}/W_{j,e}.
\end{equation}

For \verb|distribution.str="meanvar_norm"|, the normal change in mean
and variance model, the segment-specific parameter $\theta=[\mu,\sigma^2]\in\mathbb R^2$ is a vector with two values, the mean $\mu$ and variance $\sigma^2$. The Gaussian negative log likelihood is used as the loss function,
\begin{equation}
\ell_{\texttt{meanvar\_norm}}([\mu,\sigma^2],x)=[\log(2\pi\sigma^2)+(\mu-x)^2/\sigma^2]/2.
\end{equation}
For a segment from $j$ to $e$ ($1\leq j\leq e\leq N$), the variance is estimated via
\begin{equation}
  \sigma^{2*}=Y_{j,e}/W_{j,e} -\mu^*(\mu^*-2X_{j,e}/W_{j,e}).
\end{equation}
%% TODO discuss max zero var, which is the largest value that should be
%% considered numerically zero for a variance estimate. \code{INFINITY}
%% is returned for the loss if the estimated variance is smaller than max
%% zero var.

For \verb|distribution.str="poisson"|, the Poisson model, the
segment-specific parameter $\theta\geq 0$ is the rate.  The Poisson
probability mass function for a data value $x\in\{0,1,2,\dots\}$ is
\begin{equation}
  \theta^x e^{-\theta} / (x!).
\end{equation}
Maximizing the probability is equivalent to minimizing the negative
log likelihood, which simplifies to the Poisson loss,
\begin{equation}
  \ell_{\texttt{poisson}}(\theta,x)=\theta - x\log \theta.
\end{equation}

@@

\section{Algorithm} \label{sec:algorithm}
This section gives the equations for the binary segmentation algorithm.

\subsection{Min loss and difference}
The iterative algorithm involves computing the min loss over
candidate segments, each of which is defined by a start $j$ and end
$e$ position ($1<j<e<N$).

\begin{equation}
L_{j,e} = \min_{\theta} \sum_{i=j}^e w_i \ell(\theta, x_i).
\end{equation}
The minimization problem is over the model parameter $\theta$ (which affect the center and possibly scale/variance).

\paragraph{Best loss difference.}
Since different segments have different loss values before and after splitting, the algorithm must choose the next segment to split by minimizing a loss difference value.
Let $D_{j,e}(t)=L_{j,t}+L_{t+1,n}-L_{j,e}<0$ be the loss difference after splitting segment $(j,e)$ after $t$.
The $L_{j,e}$ term is the loss before the split, and $L_{j,t}+L_{t+1,n}$ is the loss after the split, which is necessarily smaller.
We define the best loss difference/split on segment $(j,e)$ as
\begin{equation}
  \label{eq:fje}
  f_{j,e}=\min,\argmin_{t\in \{j,\dots,e-1\}} D_{j,e}(t).
\end{equation}
%% We can re-write this minimization as
%% \begin{equation}
%%   f_{j,e}=\min,\argmin -L_{j,e}+ \begin{cases}
%%     D_{j,e}(j) = L_{j,j}+L_{j+1,e} \\
%%     %D_{j,e}(j+1) = L_{j,j+1}+L_{j+2,e} \\
%%     \vdots\\
%%     D_{j,e}(e-1) = L_{j,e-1}+L_{e,e} \\
%%   \end{cases}
%%   \label{eq:fje}
%% \end{equation}
\paragraph{Efficient computation.}
To compute the loss differences $D_{j,e}(j),\dots,D_{j,e}(e-1)$, we need to compute the sequence of values $L_{j,j}, \dots, L_{j,e-1}$ (loss values for possible new segments before candidate split points), and the sequence of values $L_{j+1,e}, \dots, L_{e,e}$ (loss values for possible new segments after candidate split points).
These sequences can be computed efficiently for the five loss functions that we implemented in \code{binsegRcpp}.
The time complexity of computing these sequences depends on the loss function, and the size of the segment to split, $s=e-j+1$.
For the case of negative log likelihood of Normal and Poisson distributions, the cumulative sum trick can be used to compute these sequences in linear $O(s)$ time.
For the case of the L1 loss and negative log likelihood of Laplace distribution, these sequences can be computed efficiently in log-linear $O(s\log s)$ time using a cumulative median algorithm.
To implement the cumulative median, we used a modified version of the L1 loss minimization code proposed by \citet{Drouin2017mmit}.

\subsection{Algorithm pseudo-code}

The binary segmentation algorithm needs to keep track of a set of segments that can be split, $\mathcal S_k$ for iteration $k$.
Each segment $(j,e,d,t)\in\mathcal S_k$ is represented by the start/end positions $(j,e)$ along with the corresponding min loss difference $d<0$, as well as the best split point $t\in\{j,\dots,e-1\}$.

\paragraph{Initialization.}
The initialization of the algorithm is as follows.
Let $\mathcal L_1=L_{1,n}$ be the loss with one segment,
and let $ f_{1,n}$ be the best loss difference and split point for the entire data.
We initialize the set of segments to split as $\mathcal S_1=\{(1,N,f_{1,n})\}$.

\paragraph{Recursive update rules.}
Let $S\in\{ 2, \dots, N\}$ be the user-specified max number of segments to consider.
The binary segmentation algorithm recursively computes the following quantities for each iteration $k\in\{2,\dots,S\}$.
First, the best segment to split is computed by greedy minimization of the loss decrease over all splittable segments.
\begin{equation}
  j^*_k, e^*_k, d^*_k, t^*_k = \argmin_{(j,e,d,t)\in\mathcal S_{k-1}} d.
\end{equation}
Ideally, the minimization above takes $O(1)$ constant time, if the set
of splittable segments is stored in an efficient container (for example, C++ Standard Template Library multiset or priority\_queue). As a
baseline, we also consider a linked list, which results in a
sub-optimal $O(k)$ linear time search at iteration $k$ (see
Section~\ref{sec:data-structures} for details).

@@

Third, we construct a set of two new segments to consider:
\begin{equation}
  \mathcal N_k=\{
  (j^*_k, t^*_k),
  (t^*_k+1,e^*_k)
  \}.
\end{equation}

Fourth, we subset the two new segments, to consider only the segments that could possibly be split:
\begin{equation}
  \label{eq:Vk}
  \mathcal V_k=\{(j,e,f_{j,e})\mid (j,e)\in\mathcal N_k,\, j<e\}.
\end{equation}

Finally, we update the set of splittable segments, by removing the currently split segment, and inserting the new segments that could be split:
\begin{equation}
  \mathcal S_k=
[\mathcal S_{k-1}\setminus (j^*_k, e^*_k, d^*_k, t^*_k)]
\cup\mathcal V_k.
\end{equation}
For an efficient implementation, the time complexity of these
remove/insert operations should be sub-linear. The list container
offers constant time insertion/removal; the multiset and
priority\_queue containers offer constant time removal and log-time
insertion.

@@

\section{Other features}
We used several features of C++ to implement \pkg{binsegRcpp} efficiently.
\subsection{Cross-validation}
Note that the code supports cross-validation, in the sense that the input train set can be divided into a subtrain set (used to compute model parameters) and a validation set (used to evaluate the computed model parameters). The \code{mean} and \code{var} variables are model parameters, computed using the subtrain set, whereas the other variables depend on the data (subtrain or validation set).
The \code{ERROR} parameter is an optional block of code, that is used to return an error status code, if there are unusual input data.
\subsection{Choice of container}

In binary segmentation, a segment is split in each iteration, which results in up to two new segments that must be considered to split (see Section~\ref{sec:algo} for details).
Computing the cost of splitting a segment of size $N$ is $O(N\log N)$ time for L1/Laplace losses, and $O(N)$ time for the normal/Poisson losses.
After computing the cost for two new segments, the cost of all splittable segments must be considered, in order to identify the min cost segment to split next.
We propose using three different kinds of C++ Standard Template Library containers to store and retreive the cost of segments which have previously been considered, but not yet split (Table~\ref{tab:containers}).
First, we propose using the \code{multiset} container (red-black tree) and \code{priority\_queue} container (heap), which can be used to achieve best case time complexity.
These containers keep segments sorted by their cost values, so offer constant $O(1)$ time retreival of the min cost segment to split next; insertion takes logarithmic $O(\log n)$ time in the number of segments $n$ currently stored in the container.
Second, we propose using the \code{list} container (doubly-linked list), which is used as a baseline, to demonstrate the benefits of the other containers.
The \code{list} keeps segments stored in the order in which they were inserted, so offers constant $O(1)$ time insertion of new elements, and linear $O(n)$ time retreival of the min cost segment to split, for a container with $n$ segments currently stored.

To illustrate the performence benefits of using log-time  containers  were used to efficiently , virtual classes, static
variables, function pointers, templates, macros).

\paragraph{Loss computation.} We run each algorithm on \code{N.data}
points up to \code{max.segments =
  max.changes+1}. \code{binsegRcpp::binseg} result is a list which
contains a data table with \code{max.segments} rows and column
\code{loss} that is the square loss.  \code{changepoint::cpt.mean}
result is a list of class \code{cpt.range} with method \code{logLik}
which returns the square loss of one of the models. \code{wbs::sbs}
result is a list which contains a data frame with \code{N.data-1} rows
and \code{CUSUM} and \code{min.th} columns
TODO.
\code{fpop::multiBinSeg} result is a list with element
\code{J.est}, which is a vector of \code{max.changes} square loss
decrease values.  \code{ruptures.Binseg.predict} result is a vector of
segment ends for one model size, which can be passed to
\code{sum_of_costs} method to compute the square loss.

Three packages have implemented the normal change in mean and variance
model. \code{binsegRcpp::binseg} loss values are the normal negative
log likelihood (NLL). The \code{changepoint::logLik} function returns two
times the NLL. The \code{ruptures.Binseg}
loss is related via this equation,

\begin{equation}
  \text{NLL} = (\text{rupturesLoss} + N[1+\log(2\pi)])/2
\end{equation}

\begin{table}
  \centering
  \begin{tabular}{c|cc|cc|cc}
    case/splits: & \multicolumn{2}{c}{one } & \multicolumn{2}{c}{best/equal} & \multicolumn{2}{c}{worst/unequal } \\
    operation: & insert &
                             %remove &
                                      argmin & insert & argmin & insert & argmin\\
    \hline
    list & $O(1)$ &
    % $O(1)$ &
               $O(n)$ &
    $O(S\log S)$ & $O(S)$ & $O(S)$ & $O(S)$\\
    heap/multiset & $O(\log n)$ &
    % $O(\log N)$ &
                    $O(1)$ &
                             $O(S)$ & $O(S^2)$ & $O(S)$ & $O(S)$\\
    \hline
    Find new split & \multicolumn{2}{c}{} & \multicolumn{2}{c}{$O(N\log S)$} & \multicolumn{2}{c}{$O(NS)$}
  \end{tabular}
  \caption{container properties}
  \label{tab:containers}
\end{table}

\section{Storage and retreival of segment-specific model parameters}

To simplify the discussion in this section, we use ``segment mean'' to represent arbitrary segment-specific model parameters.
For example, segment-specific model parameters can be the mean/variance of the normal distribution, or median/scale of the Laplace distribution.
A unique feature that is provided by \pkg{binsegRcpp}, but not provided by other implementations of binary segmentation, is efficient storage and retrieval of arbitrary segment mean parameters.
In contrast, \code{changepoint::cpt.mean} returns only the segment means for the selected model, and other implementations do not return any segment means (but the returned change-point variables could be used to compute estimates).

%data_vec <- c(5, 0, 20, 15, -10, 0)
For example, we consider a simple example with six data points.
<<>>=
ex6_df <- data.frame(value = c(1, -7, 8, 10, 2, 4))
ex6_binsegRcpp <- binsegRcpp::binseg(
  "mean_norm", ex6_df$value, max.segments = 4)
ex6_binsegRcpp$splits[, .(segments, end, before.mean, after.mean,
  invalidates.index, invalidates.after)]
@
The output above is a table with four rows, one for each model size.
For a max number of segments $M$, this representation of model parameters uses $M$ rows, so is asymptotically linear, $O(M)$.
The first row represents the trivial model: one segment across all data points, with \code{end=6} equal to the index of the last data point.
The model parameter of the trivial model is \code{before.mean=3}, which is the mean of all of the data.
The other rows represent the segments that result from the recursive splitting algorithm.
For example, in the second row, \code{end=2} means that the first change-point results in a split between data points 2 and 3.
The \code{before.mean = -3} is the mean up to data point 2, and \code{after.mean=6} is the mean from data point 3 to the end.
The other columns, \code{invalidates.index} and \code{invalidates.after}, indicate the previous means which are no longer valid, for models with the given split.
For example, in the second row, \code{invalidates.index=1} identifies the first row, and \code{invalidates.after=0} identifies the mean before the split (not after), which implies that \code{before.mean=3} is no longer valid, after performing the split in the second row.
In the third row, we see a new split at \code{end=4}, resulting in new parameters \code{before.mean=9} and \code{after.mean=3}, which invalidate the previous mean in row 2 after the split (\code{after.mean=6}), while the other mean in row 2 remains valid (\code{before.mean = -3}).

In general, after running binary segmentation up to $S$ segments, we
can compute a table of estimated means (one row for each segment), for
any number of segments $k\in\{1,\dots,S\}$, using the following algorithm.
First, consider only the first $k$ rows of the \code{splits} table.
Next, set values in \code{before.mean} and \code{after.mean} columns
to missing (\code{NA}), based on values in \code{invalidates.index}
and \code{invalidates.after} columns.  Finally, sort the \code{end}
values, and use the corresponding non-missing mean values.  For $k$
segments, this algorithm is $O(k \log k)$ time due to the sort.  It is
implemented in the \code{coef} method, as shown in the example code
below:

@@

To do the analogous computation with \pkg{changepoint}, we can use the code below.

@@

@@

  \section{More technical details} \label{app:technical}

>>>>

\section{Statistical model, distributions and loss functions}

We assume there is a sequence of $N$ data $x_1,\dots,x_N\in\mathbb R$,
in which we want to find abrupt changes.  For each index
$i\in\{1,\dots,N\}$, we assume $x_i\sim\mathcal D(\theta_i)$, meaning
the observed data value $x_i$ is drawn from some distribution
$\mathcal D$ with un-observed parameter $\theta_i$, typically the mean
(and perhaps variance/scale).  We would like to compute a sequence of
parameter values, $\theta_1,\dots,\theta_N$, which is piecewise
constant.  We would like to find a sequence of change-points, such
that each segment has a parameter $\theta$ which minimizes a loss
function $\ell(\theta,x)$ that depends on the distribution $\mathcal
D$. The loss functions that we consider to minimize can be interpreted
as maximizing the log likelihood of several common distributions
(Normal, Poisson, Laplace).

@@

Two models have loss functions that involve the absolute error
function, and the Laplace distribution. The probability density
function of the Laplace distribution, for data $x\in\mathbb R$, with
location parameter $\mu\in\mathbb R$, and scale parameter $b>0$, is
\begin{equation}
  \frac{1}{2b} \exp\left(
  \frac{|\mu-x|}{-b}
  \right).
\end{equation}
For \verb|distribution.str="l1"|, the segment-specific parameter
$\theta=\mu\in\mathbb R$ is the median. In this case, maximizing the
probability corresponds to minimizing the negative log likelihood,
which simplifies to the absolute error loss function,
\begin{equation}
  \ell_{\texttt{l1}}(\mu, x) = |\mu-x|.
\end{equation}

@@

%% using the cumulative median
%% For loss functions which require the cumulative median, we use the C++ code below
%% The loss function which requires a variance estimate is the Laplace negative log likelihood; the simple L1 loss is obtained without a variance estimate.
In the proposed C++ code, a \code{class absDistribution} implements
the common operations for the \texttt{l1} and \texttt{laplace} loss
functions, including parameter estimation via a cumulative median
algorithm, which is required to efficiently compute the optimal split
point on a segment.  In particular, when we search for the best split
on a segment with $s=e-j+1$ data, we need to compute the sequence of
parameter estimates $\mu^*_{j,j},\dots,\mu^*_{j,e-1}$ before the
split, and $\mu^*_{j+1,e},\dots,\mu^*_{e,e}$ after the split.
Computing these parameter estimates and corresponding loss values can
be done in $O(s\log s)$ time, using a modified version of the L1 loss
minimization code proposed by \citet{Drouin2017mmit}.

@@

There are three models which use the cumulative sum trick to find the
best split on a segment with $s$ data in $O(s)$ time.
In particular, for a sequence of $N$ data, we compute the vectors $W,X,Y\in\mathbb R^{N+1}$ in linear $O(N)$ time via:
\begin{eqnarray}
  \text{Cumulative weights: } W_0=0,\,&& \forall t\in\{1,\dots,N\},\, W_t=\sum_{i=1}^t w_i = W_{t-1}+w_i,\\
  \text{Cumulative sum: }X_0=0,\,&& \forall t\in\{1,\dots,N\},\, X_t=\sum_{i=1}^t w_i x_i=X_{t-1}+w_i x_i,\\
  \text{Cumulative squares: }Y_0=0,\,&& \forall t\in\{1,\dots,N\},\, Y_t=\sum_{i=1}^t w_i x^2_i=Y_{t-1}+w_i x_i^2.
\end{eqnarray}

@@

For \verb|distribution.str="mean_norm"|, the normal change in mean model,
the segment-specific parameter $\theta\in\mathbb R$ is the mean.
For data $x\in\mathbb R$, mean $\mu\in\mathbb R$, and variance $\sigma^2>0$, the normal probability density function is
\begin{equation}
  \frac{1}{\sqrt{2\pi\sigma^2}}
  \exp\left[
    \frac{
      (\mu-x)^2
    }{
      -2\sigma^2
    }
    \right].
\end{equation}
Maximizing the normal probability is equivalent to minimizing the negative log-likelihood, which simplifies to the square loss,
\begin{equation}
  \ell_{\texttt{mean\_norm}}(\mu, x) = (\mu-x)^2.
\end{equation}
For a segment from $j$ to $e$ ($1\leq j\leq e\leq N$), we use the notation $W_{j,e}=W_j-W_{e-1}$ for the sum of the weights over the segment, and similarly for $X_{j,e}$ and $Y_{j,e}$. Then the mean on the segment from $j$ to $e$ can be computed in constant $O(1)$ time via
\begin{equation}
  \label{eq:mu-star}
  \mu^* = X_{j,e}/W_{j,e}.
\end{equation}

For \verb|distribution.str="meanvar_norm"|, the normal change in mean
and variance model, the segment-specific parameter $\theta=[\mu,\sigma^2]\in\mathbb R^2$ is a vector with two values, the mean $\mu$ and variance $\sigma^2$. The Gaussian negative log likelihood is used as the loss function,
\begin{equation}
\ell_{\texttt{meanvar\_norm}}([\mu,\sigma^2],x)=[\log(2\pi\sigma^2)+(\mu-x)^2/\sigma^2]/2.
\end{equation}
For a segment from $j$ to $e$ ($1\leq j\leq e\leq N$), the variance estimate is computed in constant $O(1)$ time via
\begin{equation}
  \sigma^{2*}=Y_{j,e}/W_{j,e} -\mu^*(\mu^*-2X_{j,e}/W_{j,e}).
\end{equation}
%% TODO discuss max zero var, which is the largest value that should be
%% considered numerically zero for a variance estimate. \code{INFINITY}
%% is returned for the loss if the estimated variance is smaller than max
%% zero var.

For \verb|distribution.str="poisson"|, the Poisson model, the
segment-specific parameter $\theta=\mu\geq 0$ is the rate.  The Poisson
probability mass function for a data value $x\in\{0,1,2,\dots\}$ is
\begin{equation}
  \mu^x e^{-\mu} / (x!).
\end{equation}
Maximizing the probability is equivalent to minimizing the negative
log likelihood, which simplifies to the Poisson loss,
\begin{equation}
  \ell_{\texttt{poisson}}(\mu,x)=\mu - x\log \mu.
\end{equation}
The Poisson rate parameter $\mu$ is estimated via the sample mean~(\ref{eq:mu-star}).

@@

\section{Algorithm} \label{sec:algorithm}
This section gives details of the binary segmentation algorithm.

\subsection{Efficient computation of optimal split point on a segment}
The iterative algorithm involves computing the min loss over
candidate segments, each of which is defined by a start $j$ and end
$e$ position ($1<j<e<N$).

\begin{equation}
L_{j,e} = \min_{\theta} \sum_{i=j}^e w_i \ell(\theta, x_i).
\end{equation}
The minimization problem is over the model parameter $\theta$.

\paragraph{Best loss difference.}
Since different segments have different loss values before and after splitting, the algorithm must choose the next segment to split by minimizing a loss difference value.
Let $D_{j,e}(t)=L_{j,t}+L_{t+1,n}-L_{j,e}<0$ be the loss difference after splitting segment $(j,e)$ after $t$.
The $L_{j,e}$ term is the loss before the split, and $L_{j,t}+L_{t+1,n}$ is the loss after the split, which is necessarily smaller.
We define the best loss difference/split on segment $(j,e)$ as
\begin{equation}
  \label{eq:fje}
  f_{j,e}=\min,\argmin_{t\in \{j,\dots,e-1\}} D_{j,e}(t).
\end{equation}
%% We can re-write this minimization as
%% \begin{equation}
%%   f_{j,e}=\min,\argmin -L_{j,e}+ \begin{cases}
%%     D_{j,e}(j) = L_{j,j}+L_{j+1,e} \\
%%     %D_{j,e}(j+1) = L_{j,j+1}+L_{j+2,e} \\
%%     \vdots\\
%%     D_{j,e}(e-1) = L_{j,e-1}+L_{e,e} \\
%%   \end{cases}
%%   \label{eq:fje}
%% \end{equation}
\paragraph{Efficient computation of optimal split point.}
To compute the loss differences $D_{j,e}(j),\dots,D_{j,e}(e-1)$, we need to compute the sequence of values $L_{j,j}, \dots, L_{j,e-1}$ (loss values for possible new segments before candidate split points), and the sequence of values $L_{j+1,e}, \dots, L_{e,e}$ (loss values for possible new segments after candidate split points).
These sequences can be computed efficiently for the five loss functions that we implemented in \code{binsegRcpp}.
The time complexity of computing these sequences depends on the loss function, and the size of the segment to split, $s=e-j+1$.
For the case of negative log likelihood of Normal and Poisson distributions, the cumulative sum trick can be used to compute these sequences in linear $O(s)$ time.
For the case of the L1 loss and negative log likelihood of Laplace distribution, these sequences can be computed efficiently in log-linear $O(s\log s)$ time using a cumulative median algorithm.
To implement the cumulative median, we used a modified version of the L1 loss minimization code proposed by \citet{Drouin2017mmit}.

\subsection{Containers for efficient retreival of next segment to split}
\label{sec:data-structures}
The binary segmentation algorithm needs to keep track of a set of segments that can be split, $\mathcal S_k$ for iteration $k$.
Each segment $(j,e,d,t)\in\mathcal S_k$ is represented by the start/end positions $(j,e)$ along with the corresponding min loss difference $d<0$, as well as the best split point $t\in\{j,\dots,e-1\}$.

\begin{table}
  \centering
  \begin{tabular}{c|cc|cc|cc}
    Splits: & \multicolumn{2}{c|}{one split, iteration $k$} & \multicolumn{2}{c|}{$S$ equal splits} & \multicolumn{2}{c}{$S$ unequal splits} \\
    operation: & insert &
    %remove &
    argmin & insert & argmin & insert & argmin\\
    \hline
    list & $O(1)$ &
    % $O(1)$ &
    $O(k)$ &
    $O(S)$ & $O(S^2)$ &
    $O(S)$ & $O(S)$\\
    \hline
    heap/multiset & $O(\log k)$ &
    % $O(\log N)$ &
    $O(1)$ &
    $O(S\log S)$ & $O(S)$ &
    $O(S)$ & $O(S)$\\
    \hline
    Compute loss & \multicolumn{2}{c|}{---} & \multicolumn{2}{c|}{$O(N\log S)$} & \multicolumn{2}{c}{$O(NS)$}
  \end{tabular}
  \caption{
    Asymptotic time complexity for container operations (insert new segment which could be split, argmin to find best segment to split next).
    At iteration $k$, list takes $O(k)$ time, whereas heap/multiset takes $O(\log k)$ time.
    Best case time for finding $S$ splits in $N$ data is $O(N\log S)$,
    which can be achieved by heap/multiset, because insert $O(S\log S)$ time is faster.
    However, best case time can not be achieved by list, because argmin $O(S^2)$ time is slower.
  }
  \label{tab:containers}
\end{table}
In binary segmentation, a segment is split in each iteration, which results in up to two new segments that must be considered to split. % (\ref{eq:Nk}).
Computing the cost of splitting a segment of size $s$ is $O(s\log s)$ time for L1/Laplace losses, and $O(s)$ time for the normal/Poisson losses.
After computing the cost for two new segments, the cost of all splittable segments must be considered, in order to identify the min cost segment to split next.
We propose using three different kinds of C++ Standard Template Library containers to represent $\mathcal S_k$ (set of splittable segments at iteration $k$).
Each container has a different implementation of storage and retreival of the cost of segments which have previously been considered, but not yet split (Table~\ref{tab:containers}).
First, we propose using the \code{multiset} container (red-black tree) and \code{priority\_queue} container (heap), which can be used to achieve best case time complexity.
These containers keep segments sorted by their cost values, so offer constant $O(1)$ time retreival of the min cost segment to split next; insertion takes logarithmic $O(\log n)$ time in the number of segments $n$ currently stored in the container.

\paragraph{List container as baseline.}
Second, we propose using the \code{list} container (doubly-linked list), which is used as a baseline, to demonstrate the benefits of the other containers.
The \code{list} keeps segments stored in the order in which they were inserted, so offers constant $O(1)$ time insertion of new elements, and linear $O(n)$ time retreival of the min cost segment to split, for a container with $n$ segments currently stored.
In the best case run-time of binary segmentation, there are equal sized splits at each iteration, with two new segments that could be split, but only one that is split (and the other one which stays in the set $\mathcal S_k$).
Therefore the best case size of the set of splittable segments at iteration $k$ is linear, $|\mathcal S_k|=O(k)$, and the \code{list} container therefore takes $O(k)$ time to find the next segment to split.
Going up to $S$ segments, the list therefore takes $O(S^2)$ time, which can dominate the overall time for large $S$, and prevent the code from achieving the best case $O(N \log S)$ time.

\subsection{Algorithm pseudo-code}
This section gives the update rules for the binary segmentation algorithm.
\paragraph{Initialization.}
The initialization of the algorithm is as follows.
Let $\mathcal L_1=L_{1,n}$ be the loss with one segment,
and let $ f_{1,n}$ be the best loss difference and split point for the entire data.
We initialize the set of segments to split as $\mathcal S_1=\{(1,N,f_{1,n})\}$.

\paragraph{Recursive update rules.}
Let $S\in\{ 2, \dots, N\}$ be the user-specified max number of segments to consider.
The binary segmentation algorithm recursively computes the following quantities for each iteration $k\in\{2,\dots,S\}$.
First, the best segment to split is computed by greedy minimization of the loss decrease over all splittable segments.
\begin{equation}
  j^*_k, e^*_k, d^*_k, t^*_k = \argmin_{(j,e,d,t)\in\mathcal S_{k-1}} d.
\end{equation}
Ideally, the minimization above takes $O(1)$ constant time (independent of the number of segments that could be split), if the set
of splittable segments is stored in an efficient container (for example, C++ Standard Template Library multiset or priority\_queue). As a
baseline, we also consider a linked list, which results in a
sub-optimal $O(k)$ linear time search at iteration $k$ (see
Section~\ref{sec:data-structures} for details).

@@

Third, we construct a set of two new segments to consider:
\begin{equation}
  \label{eq:Nk}
  \mathcal N_k=\{
  (j^*_k, t^*_k),
  (t^*_k+1,e^*_k)
  \}.
\end{equation}
New segments in $N_k$ could be too small to split, if there is only one data point on that segment.
We therefore consider only the segments that are large enough to be split:
\begin{equation}
  \label{eq:Vk}
  \mathcal V_k=\{(j,e,f_{j,e})\mid (j,e)\in\mathcal N_k,\, j<e\}.
\end{equation}

Finally, we update the set of splittable segments, by removing the currently split segment, and inserting the new segments that could be split:
\begin{equation}
  \mathcal S_k=
[\mathcal S_{k-1}\setminus (j^*_k, e^*_k, d^*_k, t^*_k)]
\cup\mathcal V_k.
\end{equation}
For an efficient implementation, the time complexity of these
remove/insert operations should be sub-linear in the container size,
which is true of all three containers that we consider. The list container
offers constant time insertion/removal; the multiset and
priority\_queue containers offer constant time removal and log-time
insertion.

@@

\section{Demonstration and comparison of code}
In this section, we show example code for computing binary segmentation models using several software packages.
We aim to emphasize how \pkg{binsegRcpp} makes it easy to retreive segment-specific model parameters.
To simplify the discussion in this section, we use ``segment mean'' to represent arbitrary segment-specific model parameters.
For example, segment-specific model parameters can be the mean/variance of the normal distribution, or median/scale of the Laplace distribution.

A unique feature that is provided by \pkg{binsegRcpp}, but not provided by other implementations of binary segmentation, is efficient storage and retrieval of arbitrary segment mean parameters.
In contrast, \code{changepoint::cpt.mean} returns only the segment means for the selected model, and other implementations do not return any segment means (but the returned change-point variables could be used to compute estimates).

%data_vec <- c(5, 0, 20, 15, -10, 0)
For example, we consider a simple example with six data points.
<<>>=
ex6_df <- data.frame(value = c(1, -7, 8, 10, 2, 4))
@
The code below can be used to compute models up to a max of 4 segments, using the square loss (normal change in mean, constant variance).
<<>>=
ex6_binsegRcpp <- binsegRcpp::binseg(
  "mean_norm", ex6_df$value, max.segments = 4)
ex6_binsegRcpp$splits[, .(segments, end, before.mean, after.mean,
  invalidates.index, invalidates.after)]
@
The output above is a table with four rows, one for each model size.
For a max number of segments $S$, this representation of model parameters uses $S$ rows, so is asymptotically linear space, $O(S)$.
The first row represents the trivial model: one segment across all data points, with \code{end=6} equal to the index of the last data point.
The model parameter of the trivial model is \code{before.mean=3}, which is the mean of all of the data.
The other rows represent the segments that result from the recursive splitting algorithm.
For example, in the second row, \code{end=2} means that the first change-point results in a split between data points 2 and 3.
The \code{before.mean = -3} is the mean up to data point 2, and \code{after.mean=6} is the mean from data point 3 to the end.
The other columns, \code{invalidates.index} and \code{invalidates.after}, indicate the previous means which are no longer valid, for models with the given split.
For example, in the second row, \code{invalidates.index=1} identifies the first row, and \code{invalidates.after=0} identifies the mean before the split (not after), which implies that \code{before.mean=3} is no longer valid, after performing the split in the second row.
In the third row, we see a new split at \code{end=4}, resulting in new parameters \code{before.mean=9} and \code{after.mean=3}, which invalidate the previous mean in row 2 after the split (\code{after.mean=6}), while the other mean in row 2 remains valid (\code{before.mean = -3}).

In general, after running binary segmentation up to $S$ segments, the model parameters are stored in a table with $S$ rows.
For any number of segments $k\in\{1,\dots,S\}$, we can compute a table of $k$ estimated means (one row for each segment), using the following algorithm.
First, consider only the first $k$ rows of the \code{splits} table.
Next, set values in \code{before.mean} and \code{after.mean} columns
to missing (\code{NA}), based on values in columns \code{invalidates.index}
and \code{invalidates.after}.  Finally, sort the \code{end}
values, and use the corresponding non-missing mean values.  For $k$
segments, this algorithm is $O(k \log k)$ time due to the sort.  It is
implemented in the \code{coef} method, as shown in the example code
below:

@@

To attempt the analogous computation with \pkg{changepoint}, we can use the code below.

@@

We also investigated the Poisson model (Figure~\ref{fig:timings-poisson}),
and the normal change in mean and variance model
(Figure~\ref{fig:timings-meanvar_norm}), which are implemented by
\code{blockcpd::heap}.  It is clear that \code{blockcpd::heap} has the
same asymptotic slope, but much larger constant factors, when compared
with \pkg{binsegRcpp}.

@@

  \section{More empirical timings} \label{app:more-timings}
